---
title: "Introduction to Regression"
author: "Saikat Chakraborty"
format:
  html:
    code-fold: true
editor: visual
execute:
  warning: false
---

**[DISCLAIMER:]{style="color:red;"}** [Saikat is an employee of GSK. The content, views and opinios expressed in this notebook and during the class are absolutely Saikat's own, and by no way reflects the official stand of GSK on these matters.]{style="color:red;"}

## Load all the necessary packages



```{r}
#| output: false
library(MASS)
library(tidyverse)
library(broom)
library(jtools)
#library(skimr)
```

## Load the dataset

```{r}
data(birthwt)
head(birthwt)
```
## Variable Description
**Execute the following code in your computer to explore what the above variables are**
```{r}
help(birthwt)
```

**Dimension of the dataset**
```{r}
dim(birthwt)
```



```{r}
#skimr::skim(birthwt)
```

## Question we are interested in?

**Which variables are predictors of child's birth weight?**

The first step in any analysis is to visually explore the data. Let us plot child's birth wights in the dataset.

```{r}
ggplot(data = birthwt, aes(x = bwt)) +
  geom_histogram(bins = 10, color = 'black', fill = 'lightblue3') +
  labs(x = "\nBirth Weight (in grams)", y = "Count\n") +
  theme_classic() +
  theme(text = element_text(size = 15))
```

**The birth weights are given in grams. Let us change the unit of birth weight to kg."
```{r}
ggplot(data = birthwt, aes(x = bwt/1000)) +
  geom_histogram(bins = 10, color = 'black', fill = 'lightblue3') +
  labs(x = "\nBirth Weight (in kg)", y = "Count\n") +
  theme_classic() +
  theme(text = element_text(size = 15))
```

## Is there any relation between mother's age and the child's birth weight?
```{r}
ggplot(data = birthwt, aes(y = bwt/1000, x = age)) +
  geom_point(color = 'black') +
  labs(y = "Birth Weight (in kg)\n", x = "\nMother's Age") +
  geom_smooth(method = 'lm', color = 'red') +
  theme_classic() +
  theme(text = element_text(size = 15))
```

**There dosen't appear to be any relation.**

## Is there any relation between mother's weight and the child's birth weight?
```{r}
ggplot(data = birthwt, aes(y = bwt/1000, x = lwt*0.454)) +
  #The mother's weight is given in pounds. The 0.454 multiplier is for
  #converting it into kgs
  geom_point(color = 'black') +
  labs(y = "Birth Weight (in kg)\n", x = "\nMother's Weight (in kg)") +
  geom_smooth(method = 'lm', color = 'red') +
  theme_classic() +
  theme(text = element_text(size = 15))
```
**There could be some relation.**

## Let us try to quantify this trend
```{r}

#Run the model
birthwt_modat <- birthwt %>%
  mutate(bwt2 = bwt/1000) %>%
  mutate(lwt2 = lwt*0.454)
lm_simple <- lm(bwt2 ~ lwt2, data = birthwt_modat)

#Convert the model object into tidy
lm_simple_tidy <- broom::tidy(lm_simple)

#Plot the coefficients
ggplot(lm_simple_tidy, aes(y = term, x= estimate)) +
  geom_point(color = "darkred") +
  geom_errorbar(aes(xmin = estimate - 1.96*std.error, 
                    xmax = estimate + 1.96*std.error), width = 0.25,
                color = 'black') +
  labs(x = "\nCoefficient +/- CI", y = "") +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  scale_y_discrete(labels = c("Intercept", "Mother's Weight")) +
  theme_classic() +
  theme(text = element_text(size = 15))
```

## Remove the intercept
```{r}
#Plot only the coefficient for mother's weight
ggplot(lm_simple_tidy[2,], aes(y = term, x= estimate)) +
  geom_point(color = "darkred") +
  geom_errorbar(aes(xmin = estimate - 1.96*std.error, 
                    xmax = estimate + 1.96*std.error), width = 0.25,
                color = 'black') +
   geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(x = "\nCoefficient +/- CI", y = "") +
  scale_y_discrete(labels = c("Mother's Weight")) +
  theme_classic() +
  theme(text = element_text(size = 15))
```

## How good is our model?
```{r}
#Model diagnostics
for (i in 1:6){
  plot(lm_simple, i)
}
```

## Predicted vs Observed
```{r}
ggplot(birthwt_modat, aes(x = predict(lm_simple), y=bwt2)) +
  geom_point(color = "darkred") +
  geom_abline(intercept = 0, slope = 1) +
  labs(x = "\nPredicted Birth Weight", y = "Actual Birth Weight\n") +
  theme_classic() +
  theme(text = element_text(size = 15))
```

## For the sake of completeness

```{r}
summary(lm_simple)
lm_simple_tidy
```

**As you saw, we summarized our model at the last step. Only after we have explored the model considerably. Although, we observed a significant p-value, the predicted vs observed plot suggest that our model is not that good in predicting what the newborns' birth weight is going to be. Furthermore, the R-squared suggest that our model can explain only 3% of the overall variance in the children's birthweght. Therefore, we must always be careful in interpreting our models from p-values alone.** 

## We can also ask similar question for a categorical predictor
For example, is there any relation between mother's uterine irritability and child's birth weight 

```{r}
#Plot the distribution of uterine irritability
ggplot(birthwt_modat, aes(x = as.factor(ui), y = bwt2)) +
  geom_boxplot() +
  geom_dotplot(binaxis='y', stackdir='center', fill = "red", color = "black") +
  #stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
   #              geom="pointrange", color="red") +
  labs(x = "\nUterine Irritability", y = "Child's Birth Weight\n") +
  scale_x_discrete(labels = c("Absent", "Present")) +
  theme_classic() +
  theme(text = element_text(size = 15))
```

## With the mother's smoking status during pregnancy
```{r}
#Plot the distribution of uterine irritability
ggplot(birthwt_modat, aes(x = as.factor(smoke), y = bwt2)) +
  geom_boxplot() +
  geom_dotplot(binaxis='y', stackdir='center', color = "black", fill = "red") +
  #stat_summary(fun.data=mean_sdl, fun.args = list(mult=1), 
   #              geom="pointrange", color="red") +
  labs(x = "\nMother's Smoking Status", y = "Child's Birth Weight\n") +
  scale_x_discrete(labels = c("Non-smoker", "Smoker")) +
  theme_classic() +
  theme(text = element_text(size = 15))
```

## We can fit similar models
```{r}
#Step1: Fit a linear model
lm_simple2 <- lm(bwt2 ~ as.factor(smoke), data = birthwt_modat)

#Step2: Convert the model object into tidy
lm_simple2_tidy <- broom::tidy(lm_simple2)

#Step3: Plot the coefficients
ggplot(lm_simple2_tidy, aes(y = term, x= estimate)) +
  geom_point(color = "darkred") +
  geom_errorbar(aes(xmin = estimate - 1.96*std.error, 
                    xmax = estimate + 1.96*std.error), width = 0.25,
                color = 'black') +
  labs(x = "\nCoefficient +/- CI", y = "") +
  scale_y_discrete(labels = c("Intercept", "Mother's Smoking Status")) +
  theme_classic() +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  theme(text = element_text(size = 15))

#Step4: Get the diagnostic plots
for (i in 1:6){
  plot(lm_simple2, i)
}

#Step5: Get the predicted vs observed plot
ggplot(birthwt_modat, aes(x = predict(lm_simple2), y=bwt2, color = as.factor(smoke))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  labs(x = "\nPredicted Birth Weight", y = "Actual Birth Weight\n") +
  theme_classic() +
  theme(text = element_text(size = 15))

#Step6: Summarize your model
summary(lm_simple2)


```

## Sum of squares tables
```{r}
# 1. Child's weight vs mother's weight
anova(lm_simple)
#Calculate R-squared
anova(lm_simple)$`Sum Sq`[1]/sum(anova(lm_simple)$`Sum Sq`)
#Compare with regression coefficients
summary(lm_simple)

# 2. Child's weight vs Mother's smoking status
anova(lm_simple2)
#Calculate R-squared
anova(lm_simple2)$`Sum Sq`[1]/sum(anova(lm_simple2)$`Sum Sq`)
#Compare with regression coefficients
summary(lm_simple2)
```

# Homework
Repeat simple linear regression with uterine irritability (ui) as the explanatory variable and child's weight as response

#One last plot before we go
```{r}
birthwt_modat$pred <- predict(lm_simple)
birthwt_modat2 <- birthwt_modat %>%
  pivot_longer(cols = c(bwt2, pred), names_to = "Type", values_to = "Values")
ggplot(birthwt_modat2, aes(x = lwt2, y = Values, color = Type)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  theme_classic() +
  theme(legend.title = element_blank()) +
  scale_color_manual(labels = c("Observed", "Predicted"), values = c("blue", "red")) +
  theme(text = element_text(size = 18)) +
    labs(x = "\nMother's Weight", y = "Child's Weight\n")

```